Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2021-02-07 docs/00-manuscripts/ieee/ieee-manuscript.tex
improve responses to reviewer2 and add changes to manuscript

\documentclass[letterpaper, peerreview, draftcls]{IEEEtran}

@@

\begin{document}
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?}
\IEEEpeerreviewmaketitle{Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?}
%
%
% author names and IEEE memberships
% note positions of commas and non-breaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

	%\setlength{\itemsep}{-\parsep}
	\acro{AGB}{above-ground biomass}
	\acro{ALE}{accumulated local effects}
	\acro{ALS}{airborne laser scanning}
	\acro{ANN}{artificial neural network}
	\acro{AUROC}{area under the receiver operating characteristics curve}
	\acro{BRT}{boosted regression trees}
	\acro{CART}{classification and regression trees}
	\acro{CNN}{convolutional neural networks}
	\acro{CV}{cross-validation}
	\acro{DAP}{digital aerial photogrammetry}
	\acro{ENM}{environmental niche modeling}
	\acro{FPR}{false positive rate}
	\acro{FFS}{forward feature selection}
	\acro{FS}{feature selection}
	\acro{GAM}{generalized additive model}
	\acro{GBM}{gradient boosting machine}
	\acro{GLM}{generalized linear model}
	\acro{ICGC}{Institut Cartografic i Geologic de Catalunya}
	\acro{IQR}{interquartile range}
	\acro{MARS}{multivariate adaptive regression splines}
	\acro{MBO}{model-based optimization}
	\acro{MEM}{maximum entropy model}
	\acro{ML}{machine learning}
	\acro{NDII}{normalized difference infrared index}
	\acro{NIR}{near-infrared}
	\acro{NRI}{normalized ratio index}
	\acro{NDMI}{normalized difference moisture index}
	\acro{OLS}{ordinary least squares}
	\acro{LiDAR}{light detection and ranging}
	\acro{LOWESS}{locally weighted scatter plot smoothing}
	\acro{PISR}{potential incoming solar radiation}
    \acro{PC}{principal component}
	\acro{PCA}{principal component analysis}
	\acro{PDP}{partial dependence plots}
	\acro{PLS}{partial least-squares}
	\acro{POV}{proportion of variance explained}
	\acro{RBF}{radial basis function}
	\acro{RF}{random forest}
	\acro{RMSE}{root mean square error}
	\acro{RR}{ridge regression}
	\acro{RSS}{residual sum of squares}
	\acro{SAR}{synthetic aperture radar}
	\acro{SDM}{species distribution modeling}
	\acro{SMBO}{sequential-based model optimization}
	\acro{SVM}{support vector machine}
	\acro{TPR}{true positive rate}
	\acro{VI}{vegetation index}
	\acro{XGBoost}{extreme gradient boosting}
\end{acronym}
\renewcommand{\IEEEiedlistdecl}{\relax}% reset back

@@

A \ac{PCA} was done to analyse the correlation within the variables of the different feature sets (FIXME figref).

@@

To use the full potential of the hyperspectral data, all possible vegetation indices supported by the R package hsdar (89 in total) as well as all possible \ac{NRI} combinations were calculated from the reflectances.
The following NDVI-based formula was used for the NRI calculation:

@@

The choice of NDVI serving as the base formula for NRI calculation was determined by the implementation of the \texttt{nri} function in the \texttt{hsdar} R package.

@@

where \(H\) is the conditional entropy of the response variable (class) or the feature (attribute), respectively.
In order to use this method with a numeric response (percentage defoliation of trees), the variable was discretized into equal bins and treated as a class variable.
After feature rank correlations of $> 0.9$ between different bin sizes were observed in a side analysis, \texttt{\(n_{bin}\) = 10} was found to be a reasonable setting to go with.

@@

@@

In this work, an initial design of 30 randomly composed hyperparameter settings in combination with a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is a substantial reduction of the tuning budget that is required to find a setting close to the global optimization minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage (\cite{binder2020}).
For \ac{PCA}, the number of principal components was tuned instead.
%The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_{try} = p^x_{try}$.
The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_\textrm{try} = p_\textrm{sel}^t$ as a function of the number of selected features, $p_\textrm{sel}$.
It was thus tuned on a logarithmic scale by varying $t$ between 0 (i.e. $m_\textrm{try} = 1$) and 0.5 (i.e. $m_\textrm{try}=\sqrt{p_\textrm{sel}}$).
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features left after optimizing the feature percentage during tuning.

@@

In total the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods and for the L1/L2 models, 6 feature sets $\times$ 2 models.

>>>>

% \documentclass[letterpaper, onecolumn, peerreview]{IEEEtran}
\documentclass[peerreview]{IEEEtran}

\usepackage[switch]{lineno}

@@

\begin{document}
\linenumbers
%
% paper title
% Titles are generally capitalized except for words such as a, an, and, as,
% at, but, by, for, in, nor, of, on, or, the, to and up, which are usually
% not capitalized unless they are the first or last word of the title.
% Linebreaks \\ can be used within to get better formatting as desired.
% Do not put math or special symbols in the title.
\title{Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?}
\IEEEpeerreviewmaketitle{Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?}
%
%
% author names and IEEE memberships
% note positions of commas and non-breaking spaces ( ~ ) LaTeX will not break
% a structure at a ~ so this keeps an author's name from being broken across
% two lines.
% use \thanks{} to gain access to the first footnote area
% a separate \thanks must be used for each paragraph as LaTeX2e's \thanks
% was not built to handle multiple paragraphs
%

@@

	%\setlength{\itemsep}{-\parsep}
	\acro{AGB}{above-ground biomass}
	\acro{ALE}{accumulated local effects}
	\acro{ALS}{airborne laser scanning}
	\acro{ANN}{artificial neural network}
	\acro{AUROC}{area under the receiver operating characteristics curve}
	\acro{BRT}{boosted regression trees}
	\acro{CART}{classification and regression trees}
	\acro{CNN}{convolutional neural networks}
	\acro{CV}{cross-validation}
	\acro{DAP}{digital aerial photogrammetry}
	\acro{ENM}{environmental niche modeling}
	\acro{FFS}{forward feature selection}
	\acro{FPR}{false positive rate}
	\acro{FS}{feature selection}
	\acro{GAM}{generalized additive model}
	\acro{GBM}{gradient boosting machine}
	\acro{GLM}{generalized linear model}
	\acro{ICGC}{Institut Cartografic i Geologic de Catalunya}
	\acro{IQR}{interquartile range}
	\acro{LiDAR}{light detection and ranging}
	\acro{LOWESS}{locally weighted scatter plot smoothing}
	\acro{MARS}{multivariate adaptive regression splines}
	\acro{MBO}{model-based optimization}
	\acro{MEM}{maximum entropy model}
	\acro{ML}{machine learning}
	\acro{NDII}{normalized difference infrared index}
	\acro{NDMI}{normalized difference moisture index}
	\acro{NIR}{near-infrared}
	\acro{NRI}{normalized ratio index}
	\acro{OLS}{ordinary least squares}
	\acro{OMNBR}{optimized multiple narrow-band reflectance}
	\acro{PCA}{principal component analysis}
	\acro{PDP}{partial dependence plots}
	\acro{PISR}{potential incoming solar radiation}
	\acro{PLS}{partial least-squares}
	\acro{POV}{proportion of variance explained}
	\acro{RBF}{radial basis function}
	\acro{RF}{random forest}
	\acro{RMSE}{root mean square error}
	\acro{RR}{ridge regression}
	\acro{RSS}{residual sum of squares}
	\acro{SAR}{synthetic aperture radar}
	\acro{SDM}{species distribution modeling}
	\acro{SMBO}{sequential-based model optimization}
	\acro{SVM}{support vector machine}
	\acro{TPR}{true positive rate}
	\acro{VI}{vegetation index}
	\acro{XGBoost}{extreme gradient boosting}
    \acro{PC}{principal component}
\end{acronym}
\renewcommand{\IEEEiedlistdecl}{\relax}% reset back

@@

A \ac{PCA} was done to analyse the correlation within the variables of the different feature sets (\autoref{fig:pca-pov}).

@@

To use the full potential of the hyperspectral data, all possible vegetation indices supported by the R package hsdar (89 in total) as well as all possible \ac{NRI} combinations were calculated.

NRIs follow the \ac{OMNBR} concept of data-driven information extraction from narrow-band indices of hyperspectral data \cite{thenkabail2000,thenkabail2018}.
In principal any index can be used for this data-driven concept.
Due to the popularity of the NDVI index in vegetation applications and the availability of this index in the \texttt{nri()} function of the \texttt{hsdar} R package, the following formula was applied to every available band combination:

@@

@@

where \(H\) is the conditional entropy of the response variable (class or Y) or the feature (attribute or X), respectively.
$H(X)$ is Shannon's Entropy \cite{shannon1948} for a variable X and $H(X, Y)$ is a joint Shannon's Entropy for a variable X with a condition to Y.
$H(X)$ itself is defined as

\begin{equation}
H(X) = - \sum_{i=1}^{n} P(x_i)\log_bP(x_i)
\end{equation}

where $b$ is the base of the logarithm used (with $b$ commonly being set to 2).

In order to use this method with a numeric response (percentage defoliation of trees), the variable was discretized into equal bins and treated as a class variable.
After feature rank correlations of $> 0.9$ between different bin sizes were observed in a side analysis, \texttt{\(n_{bin}\) = 10} was found to be a reasonable setting to go with.

@@

In total the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods and for the L1/L2 models, 6 feature sets $\times$ 2 models.
The selected hyperparameter settings are shown in \autoref{tab:hyperparameter_limits}.
In addition the code which generated the matrix is available in the research compendium of this study (\url{10.5281/zenodo.2635403}).

@@

In this work, an initial design of 30 randomly composed hyperparameter settings in combination with a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is a substantial reduction of the tuning budget that is required to find a setting close to the global optimization minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.
Tuning ranges are shown in \autoref{tab:hyperparameter_limits}.

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage \cite{binder2020}.
For \ac{PCA}, the number of principal components was tuned instead.
%The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_{try} = p^x_{try}$.
The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_\textrm{try} = p_\textrm{sel}^t$ as a function of the number of selected features, $p_\textrm{sel}$.
It was thus tuned on a logarithmic scale by varying $t$ between 0 (i.e. $m_\textrm{try} = 1$) and 0.5 (i.e. $m_\textrm{try}=\sqrt{p_\textrm{sel}}$).
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features left after optimizing the feature percentage during tuning.

@@
